---
title: "R Notebook"
output: html_notebook
---

# Reminder on descriptive statistics

```{r}
data(iris)
X = iris$Sepal.Length
```

Let's look at the histogram of this variable:

```{r}
hist(X)
```

As I said, it is possible to tell a very different stoty with the histogram if we use a different number of breaks:

```{r}
hist(X,breaks = 100)
```

The selection of the number of slices can be performed automatically in R with a least 3 different algorithms!

It is also interesting to combine the histogram with an estimation of the pdf function made by `density`:

```{r}
hist(X,freq = FALSE)
lines(density(X),col=2,lwd=2)
```

Let's now use boxplots:

```{r}
boxplot(X)
```

and to relate the boxplot with the histogram:

```{r}
par(mfrow=c(2,1))
hist(X,freq = FALSE)
lines(density(X),col=2,lwd=2)
boxplot(X,horizontal = TRUE)
```

Regarding multivariate data, it is possible for the iris data to plot several boxplots side by side because the 4 variables are all expressed on the same unit (cm).

```{r}
boxplot(iris[,-5])
```

The alternative would be:

```{r}
par(mfrow=c(2,2))
for (j in 1:4) boxplot(iris[,j],main=paste('variable',j))
```

About scatterplot / pair plot now:

```{r}
plot(iris[,-5])
```

It is possible to add to this plot an additional categorical variable (here the species):
```{r}
plot(iris[,-5],col=as.numeric(iris$Species))
```

of course, it is possible to have a classical scatter plot when looking only at 2 variables:

```{r}
plot(iris$Sepal.Length,iris$Petal.Length,col=as.numeric(iris$Species))
```

## Multivariate numerical indicators

Mean:

```{r}
data(iris)
mu = colMeans(iris[,-5])
mu
```

Covariance and correlation matrices:
```{r}
S = cov(iris[,-5])
S

C = cor(iris[,-5])
C
```

# The learning process in supervised classification

## LDA and logistic regression

Let's first apply LDA on the iris data:
```{r}
data(iris)
X = iris[,-5]
Y = as.numeric(iris$Species)

library(MASS)
system.time(f <- lda(X,Y))
```

After the learning, it is possible to examine the output:

```{r}
f
```

Afterward, if we would like to classify a new observation, we have to call the `predict` function:

```{r}
xstar = c(6, 2.9, 5, 3)
ystar = predict(f,xstar)
ystar
```

> Exercise: evaluate with the minimal setup or V-fold cross-validation the error rate that we can expect for LDA on this classification problem.

```{r}
n = nrow(X)
learn = sample(1:n,2/3*n)
f <- lda(X[learn,],Y[learn])
ystar = predict(f,X[-learn,])

# Computer the classification error
err = sum(Y[-learn] != ystar$class) / length(ystar$class)
err
```

A better version with resampling:

```{r}
n = nrow(X)
err = c()
for (i in 1:10){
  learn = sample(1:n,2/3*n)
  f <- lda(X[learn,],Y[learn])
  ystar = predict(f,X[-learn,])

  # Computer the classification error
  err[i] = sum(Y[-learn] != ystar$class) / length(ystar$class)
}
boxplot(err)
```

## Classification with KNN

A basic call to KNN is as follows:

```{r}
data(iris)
X = iris[,-5]
Y = as.numeric(iris$Species)

library(class)
?knn

xstar = c(6, 2.9, 5, 3)
ystar = knn(X,xstar,Y,k = 5)
ystar
```

> Exercise: evaluate with a V-fold cross-validation the performance of KNN (with K=5) on the iris data set.

```{r}
library(MASS)
V = 25
n = nrow(X)
fold = rep(1:V,n/V)
err.knn5 = err.lda = rep(NA,V)
for (v in 1:V){
  learn = which(fold != v)
  test = which(fold == v)
  
  # KNN
  ystar = knn(X[learn,],X[test,],Y[learn],k = 5)
  err.knn5[v] = sum(ystar != Y[test]) / length(ystar)
  
  # LDA
  f <- lda(X[learn,],Y[learn])
  ystar = predict(f,X[test,])$class
  err.lda[v] = sum(ystar != Y[test]) / length(ystar)
}
cat(paste('> KNN5:',mean(err.knn5),'+/-',sd(err.knn5),'\n'))
cat(paste('> LDA:',mean(err.lda),'+/-',sd(err.lda),'\n'))
```

> Exercise: use V-fold cross-validation to find the appropriate value for K in KNN on the iris classification problem.

```{r}
library(MASS)
V = 25 ; K = 30
n = nrow(X)
fold = rep(1:V,n/V)
err.knn = matrix(NA,K,V)

for (v in 1:V){
  learn = which(fold != v)
  test = which(fold == v)
  for (k in 1:K){
    ystar = knn(X[learn,],X[test,],Y[learn],k)
    err.knn[k,v] = sum(ystar != Y[test]) / length(ystar)
  }
}
plot(rowMeans(err.knn),type='b')
```

On this run, we would pick K=11.

> WARNING: To compare the performance of KNN in general with LDA, we should use double CV for both picking the best K and evaluating the performance


```{r}
library(MASS)
V = 25
n = nrow(X)
fold = rep(1:V,n/V)
err.knn = err.lda = rep(NA,V)
for (v in 1:V){
  learn = which(fold != v)
  test = which(fold == v)
  
  # KNN (with an internal CV for picking K)
  X2 = X[learn,]; Y2 = Y[learn]
  K = 30; n2 = nrow(X2)
  fold2 = rep(1:V,n2/V)
  err.knn2 = matrix(NA,K,V)

  for (v2 in 1:V){ # Start of the internal CV
    learn2 = which(fold2 != v2)
    test2 = which(fold2 == v2)
    for (k in 1:K){
      ystar = knn(X2[learn2,],X2[test2,],Y2[learn2],k)
      err.knn2[k,v2] = sum(ystar != Y2[test2]) / length(ystar)
    }
  }
  Kstar = which.min(rowMeans(err.knn2)) # Choice of K* for this fold
  
  # Evaluation of the error for this fold
  ystar = knn(X[learn,],X[test,],Y[learn],k = Kstar)
  err.knn[v] = sum(ystar != Y[test]) / length(ystar) 
  
  # LDA
  f <- lda(X[learn,],Y[learn])
  ystar = predict(f,X[test,])$class
  err.lda[v] = sum(ystar != Y[test]) / length(ystar)
}
cat(paste('> KNN:',mean(err.knn),'+/-',sd(err.knn),'\n'))
cat(paste('> LDA:',mean(err.lda),'+/-',sd(err.lda),'\n'))
```

## Classification with SVM

The SVM function is provided in R within different packages. In particular, the package 'e1071' provides a nice implementation of SVM.

```{r}
#install.packages('e1071')
library(e1071)
```

A basic use of SVM on the iris data would be:

```{r}
data(iris)
X = iris[,-5]
Y = as.numeric(iris$Species)

xstar = c(6, 2.9, 5, 3)
f.svm = svm(X,Y,kernel = 'radial',gamma = 1,type = "C-classification")
ystar = predict(f.svm,matrix(xstar,nrow=1))
ystar
```

> Exercise: use cross-validation to find the best value for gamma within the RBF kernel for this classification task.

```{r}
V = 25 ; GAMMA = seq(0.01,1,by=0.01)
n = nrow(X)
fold = rep(1:V,n/V)
err.svm = matrix(NA,length(GAMMA),V)

for (v in 1:V){
  learn = which(fold != v)
  test = which(fold == v)
  for (j in 1:length(GAMMA)){
    gamma = GAMMA[j]
    f.svm = svm(X[learn,],Y[learn],kernel='radial',gamma=gamma,
                type= "C-classification")
    ystar = predict(f.svm,X[test,])
    err.svm[j,v] = sum(ystar != Y[test]) / length(ystar)
  }
}
plot(GAMMA,rowMeans(err.svm),type='b')
title(main=paste('Best gamma value:',GAMMA[which.min(rowMeans(err.svm))],'\n'))
```

> Exercise: apply double cross-validation to find the best kernel (among linear, radial and polynomial) for SVM and compare to KNN and LDA (with 10-fold CV)

```{r}
V = 10; n = nrow(X)
fold = rep(1:V,n/V)
err.knn = err.lda = rep(NA,V)
err.svm1 = err.svm2 = err.svm3 = rep(NA,V)
for (v in 1:V){
  learn = which(fold != v)
  test = which(fold == v)
  
  # KNN (with an internal CV for picking K)
  X2 = X[learn,]; Y2 = Y[learn]
  K = 15; n2 = nrow(X2)
  fold2 = rep(1:V,n2/V)
  err.knn2 = matrix(NA,K,V)

  for (v2 in 1:V){ # Start of the internal CV
    learn2 = which(fold2 != v2)
    test2 = which(fold2 == v2)
    for (k in 1:K){
      ystar = knn(X2[learn2,],X2[test2,],Y2[learn2],k)
      err.knn2[k,v2] = sum(ystar != Y2[test2]) / length(ystar)
    }
  }
  Kstar = which.min(rowMeans(err.knn2)) # Choice of K* for this fold
  
  # Evaluation of the error for this fold
  ystar = knn(X[learn,],X[test,],Y[learn],k = Kstar)
  err.knn[v] = sum(ystar != Y[test]) / length(ystar) 
  
  # LDA #############
  f <- lda(X[learn,],Y[learn])
  ystar = predict(f,X[test,])$class
  err.lda[v] = sum(ystar != Y[test]) / length(ystar)
  
  # SVM linear #############
  f.svm = svm(X[learn,],Y[learn],kernel = 'linear',type = "C-classification")
  ystar = predict(f.svm,X[test,])
  err.svm1[v] = sum(ystar != Y[test]) / length(ystar)
  
  # SVM radial #############
  X2 = X[learn,]; Y2 = Y[learn]; n2 = nrow(X2)
  GAMMA = seq(0.01,1,by=0.01); int.svm2 = matrix(NA,length(GAMMA),V)
  fold2 = rep(1:V,n2/V)

  for (v2 in 1:V){ # Start of the internal CV
    learn2 = which(fold2 != v2)
    test2 = which(fold2 == v2)
    for (j in 1:length(GAMMA)){
      gamma = GAMMA[j]
      f.svm = svm(X2[learn2,],Y2[learn2],kernel='radial',gamma=gamma,
                type= "C-classification")
      ystar2 = predict(f.svm,X2[test2,])
      int.svm2[j,v2] = sum(ystar2 != Y2[test2]) / length(ystar2)
    }
  }
  GammaStar = GAMMA[which.min(rowMeans(int.svm2))]
  
  # Evaluation of the error for this fold
  f.svm = svm(X[learn,],Y[learn],kernel = 'radial',gamma = GammaStar,
              type = "C-classification")
  ystar = predict(f.svm,X[test,])
  err.svm2[v] = sum(ystar != Y[test]) / length(ystar)
}
cat(paste('> KNN:',mean(err.knn),'+/-',sd(err.knn),'\n'))
cat(paste('> LDA:',mean(err.lda),'+/-',sd(err.lda),'\n'))
cat(paste('> SVM linear:',mean(err.svm1),'+/-',sd(err.svm1),'\n'))
cat(paste('> SVM radial:',mean(err.svm2),'+/-',sd(err.svm2),'\n'))
```


# Unsupervised learning

## Dimension reduction with PCA

```{r}
data(iris)
X = iris[,-5]
cls = as.numeric(iris$Species)

# The usual manner to do PCA in R
pca = princomp(X)
biplot(pca)
```

We can compare the observations in the orignal space with their projection on the new variables:

```{r}
Y = predict(pca)[,1:2]

# coordinates of the 1st individual in the original space (X)
X[1,]

# coordinates of the 1st individual in the projection space (Y = X * U)
Y[1,]

plot(Y,col=cls)
```

## Another example of PCA with the decatholon data

Let's first load the data (from internet):

```{r}
decathlon = read.table("http://math.agrocampus-ouest.fr/infoglueDeliverLive/digitalAssets/108286_decathlon.csv",header=TRUE,sep=';')

dim(X)
```

```{r}
X =  decathlon[,2:11]
rownames(X) = decathlon$X
```

Excercise: Perform
```{r}
# 500m has a strong correlation ← means very important 
pca = princomp(X)
par(mfrow=c(1,2))
biplot(pca, col=c(0,2))

Y = predict(pca)[,1:2]
plot(Y, type='n')
text(Y, labels=rownames(X))
```

```{r}
Y = predict(pca)[,1:2]

# coordinates of the 1st individual in the original space (X)
X[1,]

# coordinates of the 1st individual in the projection space (Y = X * U)
Y[1,]

plot(Y, type='n')
text(Y, labels=rownames(X))
```


```{r}
summary(pca)
```

Here, two PCA components are enough to explain 93% of the variance
The other rule is to look for a break in the eigenvalue script
```{r}
screeplot(pca)
```

On this example, we perhaps would prefer $d=3$
Normalization: There is no firm theory to ensure that normalization works

