
---
title: "R Notebook"
output: html_notebook
---

# Reminder on descriptive statistics
```{r}
data(iris)
X = iris$Sepal.Length
```

It tells a vert different story with the histogram if we use the different number of breaks:

```{r}
hist(X, breaks=nclass.FD(X))
```

```{r}
hist(X, breaks=40)
```
The selection of the number of slices can be performed automatically in R with at least 3 different algorithms

It is also interesting to combine the histogram with an estimation of the pdf function mode by 'density':

```{r}
hist(X, freq=FALSE)
lines(density(X), col=2, lwd=2)
```


```{r}
boxplot(X)
```

```{r}
par(mfrow=c(2,1))
hist(X, freq=FALSE)
lines(density(X), col=2, lwd=2)
boxplot(X, horizontal=TRUE)
```
Regarding multivariate data, it is possible for the iris data to plot several boxplots side by side because the 4 variables are all expressed on the same unit(cm).

```{r}
boxplot(iris[,-5])

```
The alternative would be:

```{r}
par(mfrrow=c(2,2))
for (j in 1:4) boxplot(iris[,j], main=paste("variable", j))
```

```{r}
plot(iris[,-5])
```

It is possible to add to this plot an additional species):
```{r}
plot(iris[,-5], col=as.numeric(iris$Species))
```

```{r}
plot(iris$Sepal.Length, iris$Petal.Length, col=as.numeric(iris$Species))
```

Mean of each variable 
```{r}
data(iris)
mu = colMeans(iris[, -5])
mu
# Cannot use mean function as it calculates mean of all values and returns scalar
## Multivariate numerical indicators

```

Covariance and correlation matrices:
```{r}
S = cov(iris[,-5])
S
C = cor(iris[,-5])
C
```

# The learning process in supervised classification

## LDA and logistic regression

Let's first apply LDA on the iris data:
```{r}
data(iris)
X = iris[,-5]
Y = as.numeric(iris$Species)

library(MASS)
system.time((f = lda(X,Y)))
```

After the learning, it is possible to examine the output:

```{r}
f
```
Call:
lda(X, Y)

Afterward, if we would like to classify a new observation, we have to call the `predict` fuction:

```{r}
xstar = c(6, 2.9, 5, 3) # Four variable to make one new observation
ystar = predict(f, xstar)
ystar
```
Posterior probability is the class it is classified as
→3 in this case


```{r}
f = lda(X,Y, cv=TRUE)
predict(f, xstar)
```


```{r}
n = nrow(X)
learn = sample(1:n, 2/3*n)
f <- lda(X[learn,], Y[learn])
ystar = predict(f, X[-learn,])

# Compute the classification error
err = sum(Y[-learn] != ystar$class) / length(ystar$class)
err
```
There is no error in this case, model parfectly predict the class of each observation 

```{r}
n = nrow(X)
err = c()
for (i in 1:10){
  learn = sample(1:n, 2/3*n) # This is the stochastic sampling process
  f <- lda(X[learn,], Y[learn])
  ystar = predict(f, X[-learn,])
  
  # Compute the classification error
  
  err[i] = sum(Y[-learn] != ystar$class) / length(ystar$class)
}
boxplot(err)
```
}
```